---
title: "Sinusoidal model"
author: "Adeline Leclercq Samson"
date: "2023-01-11"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description of the sinusoidal model

The observations along the tusk are denoted $Y_i$ for $i=1, \ldots, n$, with the corresponding  position along the tusk  denoted $x_i$.
The model is the following
$$Y_i = f(x_i, \varphi)+\varepsilon_i $$
with $\varepsilon_i$ a  random noise assumed to be normally distributed with mean $0$ and variance $\omega^2$.
The regression function is a periodic sinusoidal function
$$f(x, \varphi) = A \sin(g(x)+b) + B\sin(2g(x)+2b+\pi/2) $$
with function $g$ defined as
$$ g(x) = ax+\xi_x$$
and finally $\xi_x$ is assumed to be a random Ornstein-Uhlenbeck process 
$$d\xi_x = -\beta \xi_xdx+\sigma dW_x $$

The objective is to estimate the parameters $\varphi = (A,B,a,b)$; $\psi = e^{\beta\Delta}$ where $\Delta$ is the step size between two observations and $\gamma^2 = \frac{\sigma^2}{2\beta}(1-e^{-2\beta\Delta})$. 

## Simulation of trajectories

Let us start with some simulations.
```{r}
# parameters
A = 1/2
B = -1/4
b = 1
a = 0.1
beta = 0.05
sigma = 0.1
omega = 0.01

# simulations
n = 500
x = 1:n
delta = 1
psi = exp(-delta*beta)
gamma = sigma/sqrt(2*beta)*sqrt(1-psi^2)
xi = rep(0,n)
for (i in 2:n){
  xi[i] = xi[i-1]*psi + rnorm(1,0, gamma)
}
gx = a*x+xi
fx = A*sin(gx+b) + B*sin(2*gx + 2*b + pi/2)
Y = fx + rnorm(n,0,omega)
plot(x,Y,type = "l")

```


## Estimation with the EM algorithm

The EM algorithm is based on the complete log-likelihood of the model.
The solution of the hidden process $\xi_x$ is
$$\xi_{x+\Delta} = \xi_x e^{-\beta\Delta} + \int_{x}^{x+\Delta} \sigma e^{\beta(s-x)}dW_s $$
such that the transition density is
$$p(\xi_{x+\delta}|\xi_x) = \mathcal{N}(\xi_x e^{-\beta\Delta}, \frac{\sigma^2}{2\beta}(1-e^{-2\beta\Delta}))$$

The complete log-likelihood is thus
\begin{eqnarray*}
\log L(Y,\xi,\theta)&=&\sum_{i=1}^n\log p(Y_i|\xi_i) +\sum_{i=1}^n\log p(\xi_{i}|\xi_{i-1}) +\log p(\xi_1)\\
&=& -\sum_{i=1}^n \frac{(Y_i-f(x_i, \varphi))^2}{2\omega^2}-\frac{n}{2}\log(\omega^2)\\
&&- \sum_{i=1}^n\frac{(\xi_i-\xi_{i-1}e^{-\beta\Delta})^2}{\frac{\sigma^2}{\beta}(1-e^{-2\beta\Delta})} - \frac{n}{2}\log(\frac{\sigma^2}{2\beta}(1-e^{-2\beta\Delta})) \\
&=& -\sum_{i=1}^n \frac{(Y_i-f(x_i, \varphi))^2}{2\omega^2}-\frac{n}{2}\log(\omega^2)\\
&&- \sum_{i=1}^n\frac{(\xi_i-\xi_{i-1}\psi)^2}{2\gamma^2} - \frac{n}{2}\log(\gamma^2) 
\end{eqnarray*}

The EM algorithm proceeds at iteration $k$ with the two following steps, given the current value of the parameters $\theta^k$

- E step: calculation of $Q(\theta, \theta^k)$
- M step: update of the parameters $\theta^{k+1}=\arg\max_\theta Q(\theta, \theta^k)$

#### E step

The condition distribution $p(\xi|Y; \theta^k)$ is not explicit because the regression function is not linear. We should proceed with a MCMC algorithm to sample from this distribution. This will lead to a stochastic version of the EM algorithm, namely the SAEM algorithm.

#### M step

We need the sufficient statistics to update the algorithm. 

The statistics are


\begin{eqnarray*}
S_1(\xi_i)& =& \frac{1}{n}\sum_{i=1}^n (Y_i -f(x_i(\xi_i), \varphi))^2\\
S_2(\xi_i) &=& \sum_{i=1}^n \xi_{i-1}\xi_i\\
S_3(\xi_i) &=& \sum_{i=1}^n \xi_{i-1}^2\\
S_4(\xi_i) &=& \sum_{i=1}^n \xi_i^2
\end{eqnarray*}


The update of the parameters are based on these statistics. 

#### SAEM algorithm

The steps of the SAEM algorithm are

- E step: simulation of a new trajectory $\xi^k$ with a MCMC algorithm targeting $p(\xi|Y; \theta^k)$ as stationary distribution

- SA step: stochastic approximation of the sufficient statistics

\begin{eqnarray*}
s_1^k &=& s_1^{k-1} + (1-\alpha_k)(S_1(\xi^k)-s_1^{k-1}) \\
s_2^k &=& s_2^{k-1} + (1-\alpha_k)(S_2(\xi^k)-s_2^{k-1}) \\
s_3^k &=& s_3^{k-1} + (1-\alpha_k)(S_3(\xi^k)-s_3^{k-1}) \\
s_4^k &=& s_4^{k-1} + (1-\alpha_k)(S_4(\xi^k)-s_4^{k-1}) \\
\end{eqnarray*}

- M step: update of $\theta^{k}$ using the sufficient statistics $s^k$

\begin{eqnarray*}
\widehat{\varphi}^k &=& \arg\min_\varphi \sum_{i=1}^n \left(y_i - f(x_i(\xi_i^k), \varphi)\right)^2
\\
\widehat{\psi}^k &=& \frac{s_2^k}{s_3^k}\\
\widehat{\omega^{2}}^k &=& s_1^k\\
\widehat{\gamma^{2}}^k &=& \frac{1}{n}(\widehat{\psi^2}^ks_3^k-2\widehat{\psi}^ks_2^k+s_4^k)
\end{eqnarray*}


#### MCMC
On fixe le jeu de données

```{r}
xi_true = xi

```

On initialise l'algo
```{r}
xi_init = rep(0,n)
for (i in 2:n){
  xi_init[i] = xi_init[i-1]*psi + rnorm(1,0, gamma)
}


```

