---
title: "Sinusoidal model"
author: "Adeline Leclercq Samson"
date: "2023-01-11"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
set.seed(15)

library(ggplot2)
library(reshape2)
```

## Description of the sinusoidal model

The observations along the tusk are denoted $Y_i$ for $i=1, \ldots, n$, with the corresponding  position along the tusk  denoted $x_i$.
The model is the following
$$Y_i = f(x_i, \varphi)+\varepsilon_i $$
with $\varepsilon_i$ a  random noise assumed to be normally distributed with mean $0$ and variance $\omega^2$.
The regression function is a periodic sinusoidal function
$$f(x, \varphi) = A \sin(g(x)+b) + B\sin(2g(x)+2b+\pi/2) $$
with function $g$ defined as
$$ g(x) = ax+\xi_x$$
and finally $\xi_x$ is assumed to be a random Ornstein-Uhlenbeck process
$$d\xi_x = -\beta \xi_xdx+\sigma dW_x $$

The objective is to estimate the parameters $\varphi = (A,B,a,b)$; $\psi = e^{\beta\Delta}$ where $\Delta$ is the step size between two observations and $\gamma^2 = \frac{\sigma^2}{2\beta}(1-e^{-2\beta\Delta})$.

## Simulation of trajectories

Let us start with some simulations.
```{r}
source("../../R/2_tusk/utils/simulation.R")
```

```{r}
xi <- rxi()
Y <- f(xi) + rnorm(n, 0, omega)
```

```{r}
plot(A * sin(g(rep(0, n), a) + b), type = "l", lty = 2, col = "blue", ylim = c(-.5, .75))
lines(B * sin(2 * g(rep(0, n), a) + 2 * b + pi / 2), type = "l", lty = 2, col = "red")
lines(f(rep(0, n)), type = "l")
```

```{r}
plot(xi, type = "l")
```

```{r}
plot(Y, type = "l")
```

## Estimation with the EM algorithm

The EM algorithm is based on the complete log-likelihood of the model.
The solution of the hidden process $\xi_x$ is
$$\xi_{x+\Delta} = \xi_x e^{-\beta\Delta} + \int_{x}^{x+\Delta} \sigma e^{\beta(s-x)}dW_s $$
such that the transition density is
$$p(\xi_{x+\delta}|\xi_x) = \mathcal{N}(\xi_x e^{-\beta\Delta}, \frac{\sigma^2}{2\beta}(1-e^{-2\beta\Delta}))$$

The complete log-likelihood is thus
\begin{eqnarray*}
\log L(Y,\xi,\theta)&=&\sum_{i=1}^n\log p(Y_i|\xi_i) +\sum_{i=1}^n\log p(\xi_{i}|\xi_{i-1}) +\log p(\xi_1)\\
&=& -\sum_{i=1}^n \frac{(Y_i-f(x_i, \varphi))^2}{2\omega^2}-\frac{n}{2}\log(\omega^2)\\
&&- \sum_{i=1}^n\frac{(\xi_i-\xi_{i-1}e^{-\beta\Delta})^2}{\frac{\sigma^2}{\beta}(1-e^{-2\beta\Delta})} - \frac{n}{2}\log(\frac{\sigma^2}{2\beta}(1-e^{-2\beta\Delta})) \\
&=& -\sum_{i=1}^n \frac{(Y_i-f(x_i, \varphi))^2}{2\omega^2}-\frac{n}{2}\log(\omega^2)\\
&&- \sum_{i=1}^n\frac{(\xi_i-\xi_{i-1}\psi)^2}{2\gamma^2} - \frac{n}{2}\log(\gamma^2)
\end{eqnarray*}

The EM algorithm proceeds at iteration $k$ with the two following steps, given the current value of the parameters $\theta^k$

- E step: calculation of $Q(\theta, \theta^k)$
- M step: update of the parameters $\theta^{k+1}=\arg\max_\theta Q(\theta, \theta^k)$

### E step

The condition distribution $p(\xi|Y; \theta^k)$ is not explicit because the regression function is not linear. We should proceed with a MCMC algorithm to sample from this distribution. This will lead to a stochastic version of the EM algorithm, namely the SAEM algorithm.

### M step

We need the sufficient statistics to update the algorithm.

The statistics are


\begin{eqnarray*}
S_1(\xi_i)& =& \frac{1}{n}\sum_{i=1}^n (Y_i -f(x_i(\xi_i), \varphi))^2\\
S_2(\xi_i) &=& \sum_{i=1}^n \xi_{i-1}\xi_i\\
S_3(\xi_i) &=& \sum_{i=1}^n \xi_{i-1}^2\\
S_4(\xi_i) &=& \sum_{i=1}^n \xi_i^2
\end{eqnarray*}


The update of the parameters are based on these statistics.

### SAEM algorithm

The steps of the SAEM algorithm are

- E step: simulation of a new trajectory $\xi^k$ with a MCMC algorithm targeting $p(\xi|Y; \theta^k)$ as stationary distribution

- SA step: stochastic approximation of the sufficient statistics

\begin{eqnarray*}
s_1^k &=& s_1^{k-1} + (1-\alpha_k)(S_1(\xi^k)-s_1^{k-1}) \\
s_2^k &=& s_2^{k-1} + (1-\alpha_k)(S_2(\xi^k)-s_2^{k-1}) \\
s_3^k &=& s_3^{k-1} + (1-\alpha_k)(S_3(\xi^k)-s_3^{k-1}) \\
s_4^k &=& s_4^{k-1} + (1-\alpha_k)(S_4(\xi^k)-s_4^{k-1}) \\
\end{eqnarray*}

- M step: update of $\theta^{k}$ using the sufficient statistics $s^k$

\begin{eqnarray*}
\widehat{\varphi}^k &=& \arg\min_\varphi \sum_{i=1}^n \left(y_i - f(x_i(\xi_i^k), \varphi)\right)^2
\\
\widehat{\psi}^k &=& \frac{s_2^k}{s_3^k}\\
\widehat{\omega^{2}}^k &=& s_1^k\\
\widehat{\gamma^{2}}^k &=& \frac{1}{n}(\widehat{\psi^2}^ks_3^k-2\widehat{\psi}^ks_2^k+s_4^k)
\end{eqnarray*}


#### MCMC

```{r}
source("../../R/2_tusk/utils/MCMC.R")
```

```{r}
M <- 150
mcmc.obj <- mcmc.alg(Y, M)
```

**Tirage**

```{r}
plot(xi, type = "l", col = "blue",
     ylim = c(min(min(mcmc.obj$xi.c), min(xi)), max(max(mcmc.obj$xi.c), max(xi))))
lines(mcmc.obj$xi.c, type = "l", col = "red")

plot(Y, type = "l", col = "blue")
lines(f(mcmc.obj$xi.c), type = "l", col = "red")
```

```{r}
plot(mcmc.obj$acceptance.rate[mcmc.obj$n.it,])
abline(h = acceptance.target, lty = "dashed", col = "red")
```

*Ecarts*

```{r}
worst.idx <- which.max(abs(xi - mcmc.obj$xi.c))
ggplot(data = data.frame(x = 1:M), aes(x = x)) +
  geom_line(aes(y = mcmc.obj$xi[, worst.idx]), color = "blue") +
  geom_line(aes(y = mcmc.obj$xi.p[, worst.idx]), color = "red") +
  geom_hline(aes(yintercept = xi[[worst.idx]])) +
  ylab("Y")

best.idx <- which.min(abs(xi - mcmc.obj$xi.c))
ggplot(data = data.frame(x = 1:M), aes(x = x)) +
  geom_line(aes(y = mcmc.obj$xi[, best.idx]), color = "blue") +
  geom_line(aes(y = mcmc.obj$xi.p[, best.idx]), color = "red") +
  geom_hline(aes(yintercept = xi[[best.idx]])) +
  ylab("Y")

farest.idx <- which.max(abs(xi))
ggplot(data = data.frame(x = 1:M), aes(x = x)) +
  geom_line(aes(y = mcmc.obj$xi[, farest.idx]), color = "blue") +
  geom_line(aes(y = mcmc.obj$xi.p[, farest.idx]), color = "red") +
  geom_hline(aes(yintercept = xi[[farest.idx]])) +
  ylab("Y")
```

```{r}
coeff <- max(mcmc.obj$acceptance.rate[, worst.idx]) / max(mcmc.obj$delta[, worst.idx])
ggplot(data = data.frame(x = 1:M), aes(x = x)) +
  geom_line(aes(y = mcmc.obj$acceptance.rate[, worst.idx]), color = "red") +
  geom_line(aes(y = mcmc.obj$delta[, worst.idx] * coeff), color = "blue") +
  geom_hline(aes(yintercept = acceptance.target * 1.1), linetype = 2, color = "red") +
  geom_hline(aes(yintercept = acceptance.target * .9), linetype = 2, color = "red") +
  scale_y_continuous(name = "Acceptance rate", sec.axis = sec_axis(~. / coeff, name = "Delta"))

coeff <- max(mcmc.obj$acceptance.rate[, best.idx]) / max(mcmc.obj$delta[, best.idx])
ggplot(data = data.frame(x = 1:M), aes(x = x)) +
  geom_line(aes(y = mcmc.obj$acceptance.rate[, best.idx]), color = "red") +
  geom_line(aes(y = mcmc.obj$delta[, best.idx] * coeff), color = "blue") +
  geom_hline(aes(yintercept = acceptance.target * 1.1), linetype = 2, color = "red") +
  geom_hline(aes(yintercept = acceptance.target * .9), linetype = 2, color = "red") +
  scale_y_continuous(name = "Acceptance rate", sec.axis = sec_axis(~. / coeff, name = "Delta"))
```

*xi au fil des k*

```{r}
par(mfrow = c(2, 2))
for (k.ind in round(seq(1, M, length.out = 10))) {
  plot(xi, type = "l", col = "blue",
       ylim = c(min(min(mcmc.obj$xi[k.ind,]), min(xi)), max(max(mcmc.obj$xi[k.ind,]), max(xi))),
       main = paste("k=", k.ind))
  lines(mcmc.obj$xi[k.ind,], type = "l", col = "red")
}
```

*f au fil des k*

```{r}
par(mfrow = c(2, 2))
for (k.ind in round(seq(1, M, length.out = 10))) {
  plot(Y, type = "l", col = "blue",
       main = paste("k=", k.ind))
  lines(f(mcmc.obj$xi[k.ind,]), type = "l", col = "red")
}
```

**Identifiabilité**

```{r}
p1 <- rxi()
p2 <- rxi()
p3 <- rxi()
p4 <- rxi()

par(mfrow = c(2, 2))

plot(p1, type = "l", col = 1)

plot(p1, type = "l", col = 1, ylim = c(min(c(min(p1), min(p2))), max(c(max(p1), max(p2)))))
lines(p2, type = "l", col = 2)

plot(p1, type = "l", col = 1, ylim = c(min(c(min(p1), min(p3))), max(c(max(p1), max(p3)))))
lines(p3, type = "l", col = 3)

plot(p1, type = "l", col = 1, ylim = c(min(c(min(p1), min(p4))), max(c(max(p1), max(p4)))))
lines(p4, type = "l", col = 4)
```

```{r}
par(mfrow = c(2, 2))

plot(f(p1), type = "l", col = 1)

plot(f(p1), type = "l", col = 1)
lines(f(p2), type = "l", col = 2)

plot(f(p1), type = "l", col = 1)
lines(f(p3), type = "l", col = 3)

plot(f(p1), type = "l", col = 1)
lines(f(p4), type = "l", col = 4)
```

```{r}
par(mfrow = c(2, 2))
plot(Y, type = "l", col = 1, main = "goal")

plot(Y, type = "l", col = 1, main = "MCMC")
lines(f(mcmc.obj$xi.c), type = "l", col = 2)

plot(Y, type = "l", col = 1, main = "xi simulation")
lines(f(mcmc.obj$xi.c), type = "l", col = 2)
lines(f(p1), type = "l", col = 3)

plot(Y, type = "l", col = 1, main = "xi simulation")
lines(f(mcmc.obj$xi.c), type = "l", col = 2)
lines(f(p2), type = "l", col = 4)
```

#### EM

```{r}
source("../../R/2_tusk/utils/SAEM.R")
```

```{r}
saem.obj <- saem.alg(Y)
```

```{r}
plot(xi, type = "l", col = "blue",
     ylim = c(min(min(saem.obj$mcmc$xi.c), min(xi)), max(max(saem.obj$mcmc$xi.c), max(xi))))
lines(saem.obj$mcmc$xi.c, type = "l", col = "red")

plot(Y, type = "l", col = "blue")
lines(f(saem.obj$mcmc$xi.c, A.arg = saem.obj$A.c, B.arg = saem.obj$B.c, a.arg = saem.obj$a.c, b.arg = saem.obj$b.c),
      type = "l", col = "red")

plot(f(rep(0, n)), type = "l")
lines(f(rep(0, n), A.arg = saem.obj$A.c, B.arg = saem.obj$B.c, a.arg = saem.obj$a.c, b.arg = saem.obj$b.c),
      type = "l", col = "red")
```

```{r}
par(mfrow = c(2, 2))

plot(saem.obj$omega, type = 'l', col = 'blue', main = "Omega", ylim = c(min(c(saem.obj$omega, omega)), max(saem.obj$omega, omega)))
abline(h = omega, col = 'red')

plot(saem.obj$psi, type = 'l', col = 'blue', main = "Psi", ylim = c(min(c(saem.obj$psi, psi)), max(saem.obj$psi, psi)))
abline(h = psi, col = 'red')

plot(saem.obj$gamma, type = 'l', col = 'blue', main = "gamma", ylim = c(min(c(saem.obj$gamma, gamma)), max(saem.obj$gamma, gamma)))
abline(h = gamma, col = 'red')

plot(saem.obj$A, type = 'l', col = 'blue', main = "A", ylim = c(min(c(saem.obj$A, A)), max(saem.obj$A, A)))
abline(h = A, col = 'red')

plot(saem.obj$B, type = 'l', col = 'blue', main = "B", ylim = c(min(c(saem.obj$B, B)), max(saem.obj$B, B)))
abline(h = B, col = 'red')

plot(saem.obj$b, type = 'l', col = 'blue', main = "b", ylim = c(min(c(saem.obj$b, b)), max(saem.obj$b, b)))
abline(h = b, col = 'red')

plot(saem.obj$a, type = 'l', col = 'blue', main = "a", ylim = c(min(c(saem.obj$a, a)), max(saem.obj$a, a)))
abline(h = a, col = 'red')
```

**Plan d'expérience**

*Estimation of initial a*

```{r}
params.true <- c(omega = omega, psi = psi, gamma = gamma, A = A, B = B, a = a, b = b)
params.a.est <- readRDS("../../data/2_tusk/params.est.nls.a.rds")
```

```{r}
params.a.mean <- apply(params.a.est, 2, mean)
params.a.pe <- t(apply(params.a.est, 1, function (row) (params.true - row) / params.true) * 100)
params.a.mape <- apply(abs(params.a.pe), 2, mean)
params.a.se <- apply(params.a.est, 1, function (row) (params.true - row)^2)
params.a.rmse <- sqrt(apply(params.a.se, 1, mean))
```

```{r}
df <- melt(params.a.pe, varnames = c("it", "param"))
ggplot(data = df) +
  geom_boxplot(aes(x = param, y = value))
```

```{r}
params.a.mean
params.a.mape
params.a.rmse
```

*Estimation of initial a and b*

```{r}
params.true <- c(omega = omega, psi = psi, gamma = gamma, A = A, B = B, a = a, b = b)
params.a.b.est <- readRDS("../../data/2_tusk/params.est.nls.a.b.rds")
```

```{r}
params.a.b.mean <- apply(params.a.b.est, 2, mean)
params.a.b.pe <- t(apply(params.a.b.est, 1, function (row) (params.true - row) / params.true) * 100)
params.a.b.mape <- apply(abs(params.a.b.pe), 2, mean)
params.a.b.se <- apply(params.a.b.est, 1, function (row) (params.true - row)^2)
params.a.b.rmse <- sqrt(apply(params.a.b.se, 1, mean))
```

```{r}
df <- melt(params.a.b.pe, varnames = c("it", "param"))
ggplot(data = df) +
  geom_boxplot(aes(x = param, y = value))
```

```{r}
params.a.b.mean
params.a.b.mape
params.a.b.rmse
```
