---
title: "Xi"
author: "Adeline Leclercq Samson"
date: "2023-01-11"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
set.seed(15)

library(ggplot2)
library(reshape2)
```

```{r}
source("../../R/2_tusk/utils/simulation.R")
source("../../R/2_tusk/utils/APF.R")
source("../../R/2_tusk/utils/MCMC.R")
```

## Description of the sinusoidal model

The observations along the tusk are denoted $Y_i$ for $i=1, \ldots, n$, with the corresponding  position along the tusk  denoted $x_i$.
The model is the following
$$Y_i = f(x_i, \varphi)+\varepsilon_i $$
with $\varepsilon_i$ a  random noise assumed to be normally distributed with mean $0$ and variance $\omega^2$.
The regression function is a periodic sinusoidal function
$$f(x, \varphi) = A \sin(g(x)+b) + B\sin(2g(x)+2b+\pi/2) $$
with function $g$ defined as
$$ g(x) = ax+\xi_x$$
and finally $\xi_x$ is assumed to be a random Ornstein-Uhlenbeck process
$$d\xi_x = -\beta \xi_xdx+\sigma dW_x $$

The objective is to estimate the parameters $\varphi = (A,B,a,b)$; $\psi = e^{\beta\Delta}$ where $\Delta$ is the step size between two observations and $\gamma^2 = \frac{\sigma^2}{2\beta}(1-e^{-2\beta\Delta})$.

## Estimation of $\xi_x$

The solution of the hidden process $\xi_x$ is
$$\xi_{x+\Delta} = \xi_x e^{-\beta\Delta} + \int_{x}^{x+\Delta} \sigma e^{\beta(s-x)}dW_s $$
such that the transition density is
$$p(\xi_{x+\delta}|\xi_x) = \mathcal{N}(\xi_x e^{-\beta\Delta}, \frac{\sigma^2}{2\beta}(1-e^{-2\beta\Delta}))$$

The complete log-likelihood is thus
\begin{eqnarray*}
\log L(Y,\xi,\theta)&=&\sum_{i=1}^n\log p(Y_i|\xi_i) +\sum_{i=1}^n\log p(\xi_{i}|\xi_{i-1}) +\log p(\xi_1)\\
&=& -\sum_{i=1}^n \frac{(Y_i-f(x_i, \varphi))^2}{2\omega^2}-\frac{n}{2}\log(\omega^2)\\
&&- \sum_{i=1}^n\frac{(\xi_i-\xi_{i-1}e^{-\beta\Delta})^2}{\frac{\sigma^2}{\beta}(1-e^{-2\beta\Delta})} - \frac{n}{2}\log(\frac{\sigma^2}{2\beta}(1-e^{-2\beta\Delta})) \\
&=& -\sum_{i=1}^n \frac{(Y_i-f(x_i, \varphi))^2}{2\omega^2}-\frac{n}{2}\log(\omega^2)\\
&&- \sum_{i=1}^n\frac{(\xi_i-\xi_{i-1}\psi)^2}{2\gamma^2} - \frac{n}{2}\log(\gamma^2)
\end{eqnarray*}

### MCMC

```{r}
M <- 150
mcmc.obj <- mcmc.alg(Y, M)
```

**Tirage**

```{r}
plot(xi, type = "l", col = "blue",
     ylim = c(min(min(mcmc.obj$xi.c), min(xi)), max(max(mcmc.obj$xi.c), max(xi))))
lines(mcmc.obj$xi.c, type = "l", col = "red")

plot(Y, type = "l", col = "blue")
lines(f(x, mcmc.obj$xi.c), type = "l", col = "red")
```

```{r}
plot(mcmc.obj$acceptance.rate[mcmc.obj$n.it,])
abline(h = acceptance.target, lty = "dashed", col = "red")
```

*Ecarts*

```{r}
worst.idx <- which.max(abs(xi - mcmc.obj$xi.c))
ggplot(data = data.frame(x = 1:M), aes(x = x)) +
  geom_line(aes(y = mcmc.obj$xi[, worst.idx]), color = "blue") +
  geom_line(aes(y = mcmc.obj$xi.p[, worst.idx]), color = "red") +
  geom_hline(aes(yintercept = xi[[worst.idx]])) +
  ylab("Y")

best.idx <- which.min(abs(xi - mcmc.obj$xi.c))
ggplot(data = data.frame(x = 1:M), aes(x = x)) +
  geom_line(aes(y = mcmc.obj$xi[, best.idx]), color = "blue") +
  geom_line(aes(y = mcmc.obj$xi.p[, best.idx]), color = "red") +
  geom_hline(aes(yintercept = xi[[best.idx]])) +
  ylab("Y")

farest.idx <- which.max(abs(xi))
ggplot(data = data.frame(x = 1:M), aes(x = x)) +
  geom_line(aes(y = mcmc.obj$xi[, farest.idx]), color = "blue") +
  geom_line(aes(y = mcmc.obj$xi.p[, farest.idx]), color = "red") +
  geom_hline(aes(yintercept = xi[[farest.idx]])) +
  ylab("Y")
```

```{r}
coeff <- max(mcmc.obj$acceptance.rate[, worst.idx]) / max(mcmc.obj$delta[, worst.idx])
ggplot(data = data.frame(x = 1:M), aes(x = x)) +
  geom_line(aes(y = mcmc.obj$acceptance.rate[, worst.idx]), color = "red") +
  geom_line(aes(y = mcmc.obj$delta[, worst.idx] * coeff), color = "blue") +
  geom_hline(aes(yintercept = acceptance.target * 1.1), linetype = 2, color = "red") +
  geom_hline(aes(yintercept = acceptance.target * .9), linetype = 2, color = "red") +
  scale_y_continuous(name = "Acceptance rate", sec.axis = sec_axis(~. / coeff, name = "Delta"))

coeff <- max(mcmc.obj$acceptance.rate[, best.idx]) / max(mcmc.obj$delta[, best.idx])
ggplot(data = data.frame(x = 1:M), aes(x = x)) +
  geom_line(aes(y = mcmc.obj$acceptance.rate[, best.idx]), color = "red") +
  geom_line(aes(y = mcmc.obj$delta[, best.idx] * coeff), color = "blue") +
  geom_hline(aes(yintercept = acceptance.target * 1.1), linetype = 2, color = "red") +
  geom_hline(aes(yintercept = acceptance.target * .9), linetype = 2, color = "red") +
  scale_y_continuous(name = "Acceptance rate", sec.axis = sec_axis(~. / coeff, name = "Delta"))
```

*xi au fil des k*

```{r}
par(mfrow = c(2, 2))
for (k.ind in round(seq(1, M, length.out = 10))) {
  plot(xi, type = "l", col = "blue",
       ylim = c(min(min(mcmc.obj$xi[k.ind,]), min(xi)), max(max(mcmc.obj$xi[k.ind,]), max(xi))),
       main = paste("k=", k.ind))
  lines(mcmc.obj$xi[k.ind,], type = "l", col = "red")
}
```

*f au fil des k*

```{r}
par(mfrow = c(2, 2))
for (k.ind in round(seq(1, M, length.out = 10))) {
  plot(Y, type = "l", col = "blue",
       main = paste("k=", k.ind))
  lines(f(x, mcmc.obj$xi[k.ind,]), type = "l", col = "red")
}
```

**Identifiabilité**

```{r}
p1 <- rxi()
p2 <- rxi()
p3 <- rxi()
p4 <- rxi()

par(mfrow = c(2, 2))

plot(p1, type = "l", col = 1)

plot(p1, type = "l", col = 1, ylim = c(min(c(min(p1), min(p2))), max(c(max(p1), max(p2)))))
lines(p2, type = "l", col = 2)

plot(p1, type = "l", col = 1, ylim = c(min(c(min(p1), min(p3))), max(c(max(p1), max(p3)))))
lines(p3, type = "l", col = 3)

plot(p1, type = "l", col = 1, ylim = c(min(c(min(p1), min(p4))), max(c(max(p1), max(p4)))))
lines(p4, type = "l", col = 4)
```

```{r}
par(mfrow = c(2, 2))

plot(f(x, p1), type = "l", col = 1)

plot(f(x, p1), type = "l", col = 1)
lines(f(x, p2), type = "l", col = 2)

plot(f(x, p1), type = "l", col = 1)
lines(f(x, p3), type = "l", col = 3)

plot(f(x, p1), type = "l", col = 1)
lines(f(x, p4), type = "l", col = 4)
```

```{r}
par(mfrow = c(2, 2))
plot(Y, type = "l", col = 1, main = "goal")

plot(Y, type = "l", col = 1, main = "MCMC")
lines(f(x, mcmc.obj$xi.c), type = "l", col = 2)

plot(Y, type = "l", col = 1, main = "xi simulation")
lines(f(x, mcmc.obj$xi.c), type = "l", col = 2)
lines(f(x, p1), type = "l", col = 3)

plot(Y, type = "l", col = 1, main = "xi simulation")
lines(f(x, mcmc.obj$xi.c), type = "l", col = 2)
lines(f(x, p2), type = "l", col = 4)
```

#### APF

```{r}
M <- 500
apf.obj <- apf.alg(Y, M)
```

**Tirage**

```{r}
plot(xi, type = "l", col = "blue",
     ylim = c(min(min(apf.obj$xi.c), min(xi)), max(max(apf.obj$xi.c), max(xi))))
lines(apf.obj$xi.c, type = "l", col = "red")

plot(Y, type = "l", col = "blue")
lines(f(x, apf.obj$xi.c), type = "l", col = "red")
```

#### PMCMC

```{r}
M <- 50
pmcmc.obj <- pmcmc.alg(Y, M)
```

**Tirage**

```{r}
plot(xi, type = "l", col = "blue",
     ylim = c(min(min(pmcmc.obj$xi.c), min(xi)), max(max(pmcmc.obj$xi.c), max(xi))))
lines(pmcmc.obj$xi.c, type = "l", col = "red")

plot(Y, type = "l", col = "blue")
lines(f(x, pmcmc.obj$xi.c), type = "l", col = "red")
```

```{r}
plot(pmcmc.obj$acceptance.rate[pmcmc.obj$n.it,])
abline(h = acceptance.target, lty = "dashed", col = "red")
```

*Ecarts*

```{r}
worst.idx <- which.max(abs(xi - pmcmc.obj$xi.c))
ggplot(data = data.frame(x = 1:M), aes(x = x)) +
  geom_line(aes(y = pmcmc.obj$xi[, worst.idx]), color = "blue") +
  geom_line(aes(y = pmcmc.obj$xi.p[, worst.idx]), color = "red") +
  geom_hline(aes(yintercept = xi[[worst.idx]])) +
  ylab("Y")

best.idx <- which.min(abs(xi[2:n] - pmcmc.obj$xi.c[2:n]))
ggplot(data = data.frame(x = 1:M), aes(x = x)) +
  geom_line(aes(y = pmcmc.obj$xi[, best.idx]), color = "blue") +
  geom_line(aes(y = pmcmc.obj$xi.p[, best.idx]), color = "red") +
  geom_hline(aes(yintercept = xi[[best.idx]])) +
  ylab("Y")

farest.idx <- which.max(abs(xi))
ggplot(data = data.frame(x = 1:M), aes(x = x)) +
  geom_line(aes(y = pmcmc.obj$xi[, farest.idx]), color = "blue") +
  geom_line(aes(y = pmcmc.obj$xi.p[, farest.idx]), color = "red") +
  geom_hline(aes(yintercept = xi[[farest.idx]])) +
  ylab("Y")
```

```{r}
ggplot(data = data.frame(x = 1:M), aes(x = x)) +
  geom_line(aes(y = pmcmc.obj$acceptance.rate[, worst.idx]), color = "red") +
  geom_hline(aes(yintercept = acceptance.target), linetype = 2, color = "red") +
  scale_y_continuous(name = "Acceptance rate")

ggplot(data = data.frame(x = 1:M), aes(x = x)) +
  geom_line(aes(y = pmcmc.obj$acceptance.rate[, best.idx]), color = "red") +
  geom_hline(aes(yintercept = acceptance.target), linetype = 2, color = "red") +
  scale_y_continuous(name = "Acceptance rate")

ggplot(data = data.frame(x = 1:M), aes(x = x)) +
  geom_line(aes(y = pmcmc.obj$acceptance.rate[, farest.idx]), color = "red") +
  geom_hline(aes(yintercept = acceptance.target), linetype = 2, color = "red") +
  scale_y_continuous(name = "Acceptance rate")
```

*xi au fil des k*

```{r}
par(mfrow = c(2, 2))
for (k.ind in round(seq(1, M, length.out = 10))) {
  plot(xi, type = "l", col = "blue",
       ylim = c(min(min(pmcmc.obj$xi[k.ind,]), min(xi)), max(max(pmcmc.obj$xi[k.ind,]), max(xi))),
       main = paste("k=", k.ind))
  lines(pmcmc.obj$xi[k.ind,], type = "l", col = "red")
}
```

*f au fil des k*

```{r}
par(mfrow = c(2, 2))
for (k.ind in round(seq(1, M, length.out = 10))) {
  plot(Y, type = "l", col = "blue",
       main = paste("k=", k.ind))
  lines(f(x, pmcmc.obj$xi[k.ind,]), type = "l", col = "red")
}
```
